---
title: "WUP_04"
author: "Gianna Chevez"
date: "fall 2024"
output:
  html_document:
    df_print: paged
---

```{r include=FALSE,echo=FALSE}
require(knitr)
require(tidyverse)
require(tigerstats)
GSScsv <- read_csv("GSScsv.csv")
```

# Introduction

<!--In this section you explain what you are trying to show.  Where did the data come from?  What is the research or other question you are trying to answer?!-->

The data is in the file GSScsv.csv and came from the 2018 General Social Survey.  The question to answer is to resolve this hypothesis.  
$H_0$ There is no difference between the average score on the WORDSUM test and level of schooling DEGREE  
$H_a$ There is a significant difference in average score on the WORDSUM test between at least two levels of DEGREE  

# Methods

<!--Mention that you intend to use chatGPT to the greatest extent possible. Give credit to chatGPT when you use it to create code in your report. Decide on your methods:  use "variable analysis" or other appropriate descriptors.  Make sure to choose at least one graphical method and at least one numerical method.!-->

# Results

<!--Divide this section into two sub-sections:  One for your descriptive  results and one for your inferential results.!-->

## Descriptive Results

### Graphical Descriptive Results

<!--Graphical results here.  Make sure to show your code.  Provide appropriate labels for axes, giving units if possible, and provide a good title for the graph, too.  Use the graphical results to describe the patterns if any that exist in the data as focused toward the research question!-->

```{r}
basicCN <- ggplot(GSScsv,aes(x=DEGREE,y=WORDSUM))


basicCN + geom_boxplot()
```

There is a significant difference in the boxplot, as the Graduate group's first quartile is notably higher than the High School group's third quartile. Similarly, the Junior College group's first quartile exceeds the Less than High School group's third quartile. These comparisons highlight the gap in WORDSUM scores between different education levels. 


### Numerical Descriptive Results

<!--Numerical results go here. Use the numerical results to describe the patterns if any that exist in the data as focused toward the research question!-->


```{r}
favstats(WORDSUM~DEGREE,data=GSScsv)
```

The data shows significant disparities in WORDSUM scores by education level. Graduates have the highest first quartile of ten and median of eleven, while Less Than High School scores the lowest first quartile of seven and median of eight. The Graduate group's narrow interquartile range indicates consistent high performance, in contrast to the wider interquartile ranges for High School and Junior College. Graduates averaging ten point ten and Less Than High School averaging 10.10 and Less Than High School averaging 7.75. 

## Inferential Results

<!--State hypothesis clearly.  Make sure your discussion of the inferential test covers all the aspects that the test output produces, such as test statistic, p-value etc.  Make a decision about the null hypothesis, explain the assumptions on which the selected test/procedure was based, and why the chosen procedure satisfys the assumptions and is appropriate to answer the research question!-->

```{r}
model1 <- aov(WORDSUM~DEGREE,data=GSScsv)
summary.aov(model1)
```

The p-value is 0.000236, indicating that there is a statistical difference within these results. Therefore, additional testing is necessary to explore this difference further. 

```{r}
#we specify the margins starting at the bottom and proceeding clockwise - thus 8.7 is the left side margin to hold the names of the factors
par(mar = c(3, 17, 2, 2))
plot(TukeyHSD(model1, conf.level =0.98), las=2,)
```

The graph clearly illustrates a significant difference, as both the Less Than High School-Graduate and High School-Graduate comparisons yield values below 0. This indicates a statistically significant disparity relative to the other educational degrees. 

# Discussion and Conclusion

<!--Discussion and conclusion here.  If you found a relationship be sure to consider whether the relationship occurs because one of the variavbles causes the other, or whether they perhasps are related for some other reason.  Watch the chapter 6 videos from the GeorgeTown videos collection.!-->

In conclusion, the analysis of WORDSUM scores reveals significant disparities by educational attainment. The boxplot shows that the Graduate group has the highest median (11) and first quartile (Q1 of 10) with a narrow interquartile range (IQR of 1), indicating consistent high performance. In contrast, the Less Than High School group scores the lowest Q1 (7) and median (8), reflecting lower performance and variability. The Bachelor group shows a negative skew, while High School and Junior College groups have wider IQRs (5 and 3, respectively), indicating greater variability. Mean scores support these trends, with Graduates averaging 10.10 and Less Than High School averaging 7.75. The graph highlights significant differences, as both Less Than High School-Graduate and High School-Graduate comparisons fall below 0, emphasizing their lower performance. The p-value of 0.000236 indicates a statistically significant difference, suggesting the need for further testing to explore these disparities. Overall, higher educational attainment correlates with better WORDSUM scores and less variability.